Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-04-18 paper.tex

%
%
%Predictive rate-distortion describes the tradeoff between the cost of representing features of the past, and the quality of predictions in the future.
%Let $(X_t)$ be a stationary process.
%We consider feature representations $Z$ of the past $\overleftarrow{X}$ that are used to predict the future $\overrightarrow{X}$.
%There is a tradeoff between the complexity of $Z$ and the success of predicting the future:
%More complex representations of the past can reduce future prediction error -- up to a certain limit given by the entropy rate of the process. %This limit corresponds to perfect modeling, and results in Crutchfield and colleagues' notion of complexity.
%
%The complexity of the representation $Z$, or the \emph{rate}, is formalized as the mutual information between the past and the representation $Z$: $\operatorname{I}[Z, \overleftarrow{X}]$.
%Different distortion measures could be used as measures of prediction error.
%CITE maximize the entropy, or, equivalently, minimize the mutual information, between $Z$ and the future: $\operatorname{I}[Z, \overleftarrow{X}]$.
%
%Thus, for any process $(X_t)$, we obtain a rate-distortion tradeoff curve.
%The achievable region is populated by the random variables $Z$ such that $Z \rightarrow \left(\overleftarrow{X}\right) \rightarrow \left(\overrightarrow{X}\right)$ is a Markov chain.
%The points on the tradeoff curve are obtained by solving, for each $\beta \in [0,1]$, the problem
%\begin{equation}\label{eq:ib}
%	\max_{Z} \left( \operatorname{I}[\overrightarrow{X}, Z] - \beta \cdot \operatorname{I}[Z, (\overleftarrow{X})] \right)
%\end{equation}
%where the scope of the maximization is the class of all random variables $Z$ such that $Z - \left(\overleftarrow{X}\right) - \left(\overrightarrow{X}\right)$ is a Markov chain, that is, $Z$ is independent of the future given the past.
%This is exactly the Information Bottleneck \cite{tishby-information-1999}, applied to the past and future of a stochastic process.
%
%

@@

As in OCE (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures:
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z} \left( \operatorname{I}[X_{-M\dots -1}, Z] - \lambda \cdot \operatorname{I}[Z, X_{0\dots M-1}] \right)
\end{equation}
where $Z$ ranges over all random variables such that the Markov condition~(\ref{eq:markov}) holds:
\begin{equation}
    Z\bot X_{0\dots M-1} | X_{-M\dots -1}
\end{equation}

@@

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:ib-oce-repeat}) in a way amenable to using function approximators.
Note that the predictive rate-distortion objective~(\ref{eq:ib-oce-repeat}) can equivalently be written as
\begin{equation}\label{eq:prd-recall}
	\min_{Z : Z \bot X_{0\dots M-1} | X_{-M\dots-1}}	\left[\operatorname{H}[X_{0\dots M-1} | Z] + \lambda \operatorname{I}[ X_{-M\dots-1}, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
By the two Propositions, this objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] + \lambda \cdot \operatorname{D_{KL}}\infdivx{P_\phi(\cdot|X_{-M\dots-1})}{q}\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-recall}) by setting $Z$ to be the code generated by the encoder $\phi$.
The second proposition guarantees that, for a solution to  (\ref{eq:prd-recall}), the rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] =   \mathbb{E}_{X_{-M\dots-1}}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots-1})}{q(Z)}\right]
\end{equation}
and the prediction error is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[X_{0\dots M-1}|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] %H[X_{0\dots M-1}|X_{-M\dots-1}] % \operatorname{H}[X_{0\dots M-1}]
\end{equation}
Note that the right-hand-sides of (\ref{eq:eq1-new}-\ref{eq:eq2-new}) can both be estimated efficiently using Monte Carlo samples from $Z \sim P_\phi(X_{-M\dots-1})$.

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and prediction error for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[X_{0\dots M-1}|X_{-M\dots-1}]
  \operatorname{H}[X_{0\dots M-1}|Z] \leq   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] % \operatorname{H}[X_{0\dots M-1}]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and -- if the family of approximators is sufficiently rich---can come close to turning these into the equalities  (\ref{eq:eq1-new}-\ref{eq:eq2-new}).

@@

%\paragraph{Evaluation}
After optimizing the parameters on a set of observed trajectories, we estimate rate and prediction error on the test set.
Given parameters for $\phi, \psi, q$, we evaluate the PRD objective~(\ref{eq:bound}), rate (\ref{eq:bound1-new}), and the prediction error (\ref{eq:bound2-new}) on the test set by taking, for each time series $X_{-M}...X_{M-1}$ from the test set, a single sample $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing Monte Carlo estimates for rate
\begin{equation}\label{eq:bound-mc-rate}
\mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right] \approx	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}  \log \frac{p_{\mathcal{N}(\mu, \sigma^2)}(z)}{q(z)}
\end{equation}
where $p_{\mathcal{N}(\mu, \sigma^2)}(z)$ is the Gaussian density with $\mu, \sigma^2$ computed from $X_{-M\dots -1}$ as in (\ref{eq:z-gauss}), and prediction error 
\begin{equation}\label{eq:bound-mc-distortion}
 -	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] \approx - \frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P_\psi(X_{0\dots M-1} | Z)
\end{equation}
%and the PRD objective
%\begin{equation}\label{eq:bound-mc}
%	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
%\end{equation}
%Note that one can go from distortion to predictiveness by estimating
Thanks to (\ref{eq:bound1-new}-\ref{eq:bound2-new}), these estimates are guaranteed to be \emph{upper bounds} on the true rate and prediction error achieved by the code $Z  \sim \mathcal{N}(\mu, \sigma^2)$, up to sampling error introduced into the Monte Carlo estimation by sampling $z$ and the finite size of the test set.

@@

It is important to note that this sampling error is different from the overfitting issue affecting OCE: 
(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}) provide \emph{unbiased} estimators upper bounds, whereas overfitting \emph{biases} the values obtained by OCE.
Given that NPRD \emph{provably} provides \emph{upper bounds} on rate and prediction error (up to sampling error), one can objectively compare the quality of different estimation methods:
Among methods that provide upper bounds, the one that provides the lowest such bound for a given problem is the one giving results closest to the true curve.

We estimate predictiveness $\operatorname{I}[Z, X_{0\dots M-1}]$ with the following method.
%by estimating $\operatorname{H}[X_{0\dots M-1}]$ 
Together with $\psi$, we also create an LSTM network that estimates the marginal probability of future observation sequences $P(X_{0\dots M-1})$. Let $P_\eta$ this estimated distribution; we then estimate the predictiveness $\operatorname{I}[Z, X_{0\dots M-1}] = \operatorname{H}[X_{0\dots M-1}] - \operatorname{H}[X_{0\dots M-1}|Z]$ by the difference between the corresponding cross-entropies on the test set~(\cite{mcallester2018formal}):
%$\operatorname{H}[X_{0\dots M-1}]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{-M\dotsM-1}}\left[\log P_\eta(X_{0\dots M-1}) - P_\psi(X_{0\dots M-1} | Z)\right]
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

@@

\paragraph{OCE}
As discussed in Section~\ref{sec:oce}, OCE overestimates structure in the process due to overfitting, %resulting in biased estimates of rate and distortion.More specifically, 
overfitting will result in an overly optimistic estimate, systematically overestimating the predictiveness achieved at a given rate.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and distortion on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|X_{-M\dots -1})$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction error.
This method not only enables fair comparison between OCE and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{ineq2}:
\begin{equation}\label{eq:rate-oce}
\frac{1}{N} \sum_{X_{-M}\dots X_{-1} \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|X_{-M}\dots X_{-1})}{s(Z)}
\end{equation}
where $P(Z|X_{-M}\dots X_{-1})$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-oce}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction error on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{ineq1}:
\begin{equation}
\frac{1}{N} \sum_{X_{-M}\dots X_{M-1} \in \text{Test Data}} \E_{Z \sim P(\cdot|X_{-M \dots -1})} \log P(X_{0\dots M-1}|Z).
\end{equation}
where $P(X_{0\dots M-1}|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to (\ref{ineq1}-\ref{ineq2}), these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

Does NPRD recover this picture?
We applied PCA to samples from $Z$ computed at two different values of $\lambda$, $\lambda = 0.25$ and $\lambda = 0.6$, corresponding to the two points marked in the rate-distortion figure (\textbf{TODO do this}).
The first two principal components are shown in Figure~\ref{fig:latent}.
Samples are colored by the causal states corresponding to the pasts of the trajectories that were encoded into the respective points by NPRD.
On the left, obtained at $\lambda=0.6$, the states A and B are collapsed, as expected.
On the right, obtained at $\lambda=0.25$, the three causal states are reflected as distinct modes of $Z$.
Note that, at finite $M$, a fraction of pasts is ambiguous between the green and blue causal states; these are colored in black and NPRD maps them into a region between the modes corresponding to these states.

@@

We applied NPRD to the problem of predicting English on the level of part-of-speech tags in Section~\ref{sec:pos}.
%We verified that NPRD matches OCE in the low-rate regime where OCE is still applicable, and we fitted 
We found that the resulting curves were described well by Equation~\ref{eq:derivatives}.
%\begin{equation}\label{eq:derivatives}
%	\partial P/\partial R = \exp\left(-\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right). %&\ \ \ \partial R/\partial P = \exp\left(\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right)
%\end{equation}
We now consider the more realistic problem of prediction at the level of words, using data from multiple languages.
This problem is much closer to the task faced by a human in the process of comprehending text, having to encode prior observations so as to minimize prediction error on the upcoming words.
We will examine whether Equation~\ref{eq:derivatives} describes the resulting tradeoff in this more realistic setting, and whether it holds across languages.

@@

\paragraph{Implication 1: Less information locality requires more memory at the same level of prediction error}
Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
Then, when fixing $J$, the lower bound on surprisal provided by the proposition is higher in the case of $Y$ than in the case of $X$.
While this does not mean that every such $X$ will require less memory than every such $Y$ (since this is only a lower bound), it does entail an asymptotic version of this:
If we fix $X$ and decrease the information locality of $Y$ indefinitely strongly, $Y$ will ultimately require more memory than $X$, at the same level of experienced surprisal.

>>>>

%
%
%Predictive rate-distortion describes the tradeoff between the cost of representing features of the past, and the quality of predictions in the future.
%Let $(X_t)$ be a stationary process.
%We consider feature representations $Z$ of the past $\overleftarrow{X}$ that are used to predict the future $\overrightarrow{X}$.
%There is a tradeoff between the complexity of $Z$ and the success of predicting the future:
%More complex representations of the past can reduce future prediction loss -- up to a certain limit given by the entropy rate of the process. %This limit corresponds to perfect modeling, and results in Crutchfield and colleagues' notion of complexity.
%
%The complexity of the representation $Z$, or the \emph{rate}, is formalized as the mutual information between the past and the representation $Z$: $\operatorname{I}[Z, \overleftarrow{X}]$.
%Different distortion measures could be used as measures of prediction loss.
%CITE maximize the entropy, or, equivalently, minimize the mutual information, between $Z$ and the future: $\operatorname{I}[Z, \overleftarrow{X}]$.
%
%Thus, for any process $(X_t)$, we obtain a rate-distortion tradeoff curve.
%The achievable region is populated by the random variables $Z$ such that $Z \rightarrow \left(\overleftarrow{X}\right) \rightarrow \left(\overrightarrow{X}\right)$ is a Markov chain.
%The points on the tradeoff curve are obtained by solving, for each $\beta \in [0,1]$, the problem
%\begin{equation}\label{eq:ib}
%	\max_{Z} \left( \operatorname{I}[\overrightarrow{X}, Z] - \beta \cdot \operatorname{I}[Z, (\overleftarrow{X})] \right)
%\end{equation}
%where the scope of the maximization is the class of all random variables $Z$ such that $Z - \left(\overleftarrow{X}\right) - \left(\overrightarrow{X}\right)$ is a Markov chain, that is, $Z$ is independent of the future given the past.
%This is exactly the Information Bottleneck \cite{tishby-information-1999}, applied to the past and future of a stochastic process.
%
%

@@

As in OCE (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures:
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot X_{0\dots M-1} | X_{-M\dots-1}} \left( \operatorname{I}[Z, X_{0\dots M-1}] - \lambda \cdot \operatorname{I}[X_{-M\dots -1}, Z] \right)
\end{equation}
%where $Z$ ranges over all random variables such that the Markov condition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot X_{0\dots M-1} | X_{-M\dots -1}
%\end{equation}
It will be convenient to equivalently rewrite this as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot X_{0\dots M-1} | X_{-M\dots-1}}	\left[\operatorname{H}[X_{0\dots M-1} | Z] + \lambda \cdot\operatorname{I}[ X_{-M\dots-1}, Z]\right]
\end{equation}
where $\operatorname{H}[X_{0\dots M-1} | Z]$ is the \emph{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[X_{0\dots M-1} | Z]$. % is the reduction in prediction loss when predicting

@@

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:ib-oce-repeat}) in a way amenable to using function approximators.
Recall the predictive rate-distortion objective~(\ref{eq:prd-loss})
\begin{equation}\label{eq:prd-recall}
	\min_{Z : Z \bot X_{0\dots M-1} | X_{-M\dots-1}}	\left[\operatorname{H}[X_{0\dots M-1} | Z] + \lambda \operatorname{I}[ X_{-M\dots-1}, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
By the two Propositions, this objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] + \lambda \cdot \operatorname{D_{KL}}\infdivx{P_\phi(\cdot|X_{-M\dots-1})}{q}\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-recall}) by setting $Z$ to be the code generated by the encoder $\phi$.
The second proposition guarantees that, for a solution to  (\ref{eq:prd-recall}), the rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] =   \mathbb{E}_{X_{-M\dots-1}}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots-1})}{q(Z)}\right]
\end{equation}
and the prediction loss is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[X_{0\dots M-1}|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] %H[X_{0\dots M-1}|X_{-M\dots-1}] % \operatorname{H}[X_{0\dots M-1}]
\end{equation}
Note that the right-hand-sides of (\ref{eq:eq1-new}-\ref{eq:eq2-new}) can both be estimated efficiently using Monte Carlo samples from $Z \sim P_\phi(X_{-M\dots-1})$.

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and prediction loss for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[X_{0\dots M-1}|X_{-M\dots-1}]
  \operatorname{H}[X_{0\dots M-1}|Z] \leq   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] % \operatorname{H}[X_{0\dots M-1}]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and -- if the family of approximators is sufficiently rich---can come close to turning these into the equalities  (\ref{eq:eq1-new}-\ref{eq:eq2-new}).

@@

%\paragraph{Evaluation}
After optimizing the parameters on a set of observed trajectories, we estimate rate and prediction loss on the test set.
Given parameters for $\phi, \psi, q$, we evaluate the PRD objective~(\ref{eq:bound}), rate (\ref{eq:bound1-new}), and the prediction loss (\ref{eq:bound2-new}) on the test set by taking, for each time series $X_{-M}...X_{M-1}$ from the test set, a single sample $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing Monte Carlo estimates for rate
\begin{equation}\label{eq:bound-mc-rate}
\mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right] \approx	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}  \log \frac{p_{\mathcal{N}(\mu, \sigma^2)}(z)}{q(z)}
\end{equation}
where $p_{\mathcal{N}(\mu, \sigma^2)}(z)$ is the Gaussian density with $\mu, \sigma^2$ computed from $X_{-M\dots -1}$ as in (\ref{eq:z-gauss}), and prediction loss 
\begin{equation}\label{eq:bound-mc-distortion}
 -	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] \approx - \frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P_\psi(X_{0\dots M-1} | Z)
\end{equation}
%and the PRD objective
%\begin{equation}\label{eq:bound-mc}
%	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
%\end{equation}
%Note that one can go from distortion to predictiveness by estimating
Thanks to (\ref{eq:bound1-new}-\ref{eq:bound2-new}), these estimates are guaranteed to be \emph{upper bounds} on the true rate and prediction loss achieved by the code $Z  \sim \mathcal{N}(\mu, \sigma^2)$, up to sampling error introduced into the Monte Carlo estimation by sampling $z$ and the finite size of the test set.

@@

It is important to note that this sampling error is different from the overfitting issue affecting OCE: 
(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}) provide \emph{unbiased} estimators upper bounds, whereas overfitting \emph{biases} the values obtained by OCE.
Given that NPRD \emph{provably} provides \emph{upper bounds} on rate and prediction loss (up to sampling error), one can objectively compare the quality of different estimation methods:
Among methods that provide upper bounds, the one that provides the lowest such bound for a given problem is the one giving results closest to the true curve.

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, X_{0\dots M-1}]$ with the following method.
%by estimating $\operatorname{H}[X_{0\dots M-1}]$ 
Together with $\psi$, we also create an LSTM network that estimates the marginal probability of future observation sequences $P(X_{0\dots M-1})$. Let $P_\eta$ this estimated distribution; we then estimate the predictiveness $\operatorname{I}[Z, X_{0\dots M-1}] = \operatorname{H}[X_{0\dots M-1}] - \operatorname{H}[X_{0\dots M-1}|Z]$ by the difference between the corresponding cross-entropies on the test set~(\cite{mcallester2018formal}):
%$\operatorname{H}[X_{0\dots M-1}]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{-M\dotsM-1}}\left[\log P_\eta(X_{0\dots M-1}) - P_\psi(X_{0\dots M-1} | Z)\right]
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

@@

\paragraph{OCE}
As discussed in Section~\ref{sec:oce}, OCE is affected by overfitting, %overestimates structure in the process due to overfitting, %resulting in biased estimates of rate and distortion.More specifically, 
%overfitting will result in an overly optimistic estimate, 
and will systematically overestimate the predictiveness achieved at a given rate.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and predictiveness on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|X_{-M\dots -1})$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction loss.
This method not only enables fair comparison between OCE and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{ineq2}:
\begin{equation}\label{eq:rate-oce}
\frac{1}{N} \sum_{X_{-M}\dots X_{-1} \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|X_{-M}\dots X_{-1})}{s(Z)}
\end{equation}
where $P(Z|X_{-M}\dots X_{-1})$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-oce}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{ineq1}:
\begin{equation}
\frac{1}{N} \sum_{X_{-M}\dots X_{M-1} \in \text{Test Data}} \E_{Z \sim P(\cdot|X_{-M \dots -1})} \log P(X_{0\dots M-1}|Z).
\end{equation}
where $P(X_{0\dots M-1}|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to (\ref{ineq1}-\ref{ineq2}), these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

Does NPRD recover this picture?
We applied PCA to samples from $Z$ computed at two different values of $\lambda$, $\lambda = 0.25$ and $\lambda = 0.6$. % corresponding to the two points marked in the rate-distortion figure (\textbf{TODO do this}).
The first two principal components are shown in Figure~\ref{fig:latent}.
Samples are colored by the causal states corresponding to the pasts of the trajectories that were encoded into the respective points by NPRD.
On the left, obtained at $\lambda=0.6$, the states A and B are collapsed, as expected.
On the right, obtained at $\lambda=0.25$, the three causal states are reflected as distinct modes of $Z$.
Note that, at finite $M$, a fraction of pasts is ambiguous between the green and blue causal states; these are colored in black and NPRD maps them into a region between the modes corresponding to these states.

@@

We applied NPRD to the problem of predicting English on the level of part-of-speech tags in Section~\ref{sec:pos}.
%We verified that NPRD matches OCE in the low-rate regime where OCE is still applicable, and we fitted 
We found that the resulting curves were described well by Equation~\ref{eq:derivatives}.
%\begin{equation}\label{eq:derivatives}
%	\partial P/\partial R = \exp\left(-\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right). %&\ \ \ \partial R/\partial P = \exp\left(\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right)
%\end{equation}
We now consider the more realistic problem of prediction at the level of words, using data from multiple languages.
This problem is much closer to the task faced by a human in the process of comprehending text, having to encode prior observations so as to minimize prediction loss on the upcoming words.
We will examine whether Equation~\ref{eq:derivatives} describes the resulting tradeoff in this more realistic setting, and whether it holds across languages.

@@

\paragraph{Implication 1: Less information locality requires more memory at the same level of prediction loss}
Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
Then, when fixing $J$, the lower bound on surprisal provided by the proposition is higher in the case of $Y$ than in the case of $X$.
While this does not mean that every such $X$ will require less memory than every such $Y$ (since this is only a lower bound), it does entail an asymptotic version of this:
If we fix $X$ and decrease the information locality of $Y$ indefinitely strongly, $Y$ will ultimately require more memory than $X$, at the same level of experienced surprisal.
